---
title: "Comparison to metadata extraction with oro.dicom and oro.nifti"
author: "Pamela Russell"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Comparison with oro packages}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
rm(list=ls())
suppressPackageStartupMessages(library(knitr))
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE
)
```


```{r suppress_eval, echo = FALSE}
# Check if this is running locally and if not, suppress evaluation
if(!utils::file_test("-f", "~/Dropbox/radtools_vignette_data/prostate/000008.dcm")) {
  knitr::opts_chunk$set(eval = FALSE)
  message("Note: code examples will not be evaluated because they depend on local data.")
}
```

# Overview

Radtools forwards to [oro.dicom](https://cran.r-project.org/package=oro.dicom) and [oro.nifti](https://cran.r-project.org/package=oro.nifti) under the hood for reading images and extracting data from them. The added value of radtools is to provide simple, uniform functions to access metadata in convenient formats, reducing the need for code duplication or for understanding the finer points of oro.dicom and oro.nifti. In this vignette we compare some of radtools' functionality for metadata extraction to equivalent code using oro.dicom and oro.nifti.

Load sample data:

```{r import_package, warning = FALSE, message = FALSE}
dicom_data <- radtools::read_dicom("~/Dropbox/radtools_vignette_data/prostate/")
nifti_data_rad <- radtools::read_nifti1("~/Dropbox/radtools_vignette_data/filtered_func_data.nii.gz")
nifti_data_oro <- oro.nifti::readNIfTI("~/Dropbox/radtools_vignette_data/filtered_func_data.nii.gz")
```


# Extracting image metadata

## Get image dimensions and number of slices

#### radtools

The functions `radtools::img_dimensions` and `radtools::num_slices` work for both DICOM and NIfTI images.

```{r img_dim_rad}
radtools::img_dimensions(dicom_data)
radtools::img_dimensions(nifti_data_rad)
radtools::num_slices(dicom_data)
radtools::num_slices(nifti_data_rad)
```

#### oro

```{r img_dim_oro}
mat_dicom <- oro.dicom::create3D(dicom_data)
dim(mat_dicom)
dim(nifti_data_oro)
dim(mat_dicom)[3] # Number of slices
dim(nifti_data_oro)[3] # Number of slices
```



## Get the actual metadata attributes present in a dataset

The `radtools::header_fields` function works for both DICOM and NIfTI data.

### DICOM

#### radtools

```{r header_fields_rad}
fields <- radtools::header_fields(dicom_data)
head(fields, 10)
```

#### oro.dicom

```{r header_field_oro}
tab <- oro.dicom::dicomTable(dicom_data$hdr)
fields <- colnames(tab)
head(fields, 10)
```


### NIfTI

All NIfTI datasets have the same metadata attributes.

#### radtools

```{r header_fields_rad_n}
fields <- radtools::header_fields(nifti_data_rad)
head(fields, 10)
```

#### oro.nifti

Not provided; functions for each individual metadata attribute are provided.


## Get DICOM metadata attribute across slices as a list

#### radtools

```{r header_value_rad}
radtools::header_value(dicom_data, "SliceLocation")
```

#### oro.dicom

```{r header_value_oro}
oro.dicom::extractHeader(dicom_data$hdr, "SliceLocation")
```


## Get the metadata for entire dataset as a table (DICOM) or list (NIFTI)

### DICOM

#### radtools

```{r dicom_header_as_mat_rad, warning = FALSE}
mat <- radtools::dicom_header_as_matrix(dicom_data)
kable(mat[1:10, 1:6])
```

#### oro.dicom

```{r dicom_header_as_mat_oro}
mat <- oro.dicom::dicomTable(dicom_data$hdr)
kable(mat[1:10, 1:6])
```

### NIfTI

#### radtools

```{r}
nifti_header_vals <- radtools::nifti1_header_values(nifti_data_rad)
head(nifti_header_vals[names(nifti_header_vals) != ".Data"])
```

#### oro.nifti

Functionality not available in oro.nifti


## Get constant (across slices) DICOM attributes and their values

#### radtools

```{r dicom_const_rad}
const_attributes <- radtools::dicom_constant_header_values(dicom_data)
head(const_attributes)
```

#### oro.dicom

```{r dicom_const_oro}
tab <- oro.dicom::dicomTable(dicom_data$hdr)
const_cols <- apply(tab, 2, function(x) {length(unique(x)) == 1})
const_attributes <- as.list(tab[1, const_cols])
head(const_attributes)
```


# Extracting image data

#### radtools

The `radtools::img_data_to_mat` function works for both DICOM and NIfTI data.

```{r img_data_to_mat_rad}
mat_dicom <- radtools::img_data_to_mat(dicom_data)
dim(mat_dicom)
mat_nifti <- radtools::img_data_to_mat(nifti_data_rad)
dim(mat_nifti)
```

#### oro

```{r img_data_to_mat_oro}
mat_dicom <- oro.dicom::create3D(dicom_data)
dim(mat_dicom)
mat_nifti <- oro.nifti::img_data(nifti_data_oro)
dim(mat_nifti)
```


# Viewing images

## DICOM

#### radtools

```{r view_slice_rad, warning = FALSE, fig.width = 5, fig.height = 5}
radtools::view_slice(dicom_data, slice = 10)
```

#### oro.dicom

```{r view_slice_oro, warning = FALSE, fig.width = 5, fig.height = 5}
mat <- oro.dicom::create3D(dicom_data)
m <- mat[,,10]
col <- grDevices::grey(0:64/64)
graphics::image(x = 1:nrow(m), y = 1:ncol(m), z = m, col = col, ann = FALSE)
```

## NIfTI

#### radtools

```{r, fig.width = 5, fig.height = 5}
mat <- radtools::img_data_to_3D_mat(nifti_data_rad, coord_extra_dim = 90)
radtools::view_slice_mat(mat, slice = 10)
```

#### oro.nifti

```{r, fig.width = 5, fig.height = 5}
oro.nifti::image(nifti_data_oro, z = 10, w = 90, plot.type = "single")
```


# DICOM standard

## radtools

#### Get the DICOM standard version used

```{r dicom_standard_version}
radtools::dicom_standard_version()
radtools::dicom_standard_web()
radtools::dicom_standard_timestamp()
```

#### Get lists of all valid attributes

Tags:

```{r tags}
tags <- radtools::dicom_all_valid_header_tags()
head(tags, 10)
```

Names:

```{r names}
names <- radtools::dicom_all_valid_header_names()
head(names, 10)
```

Keywords:

```{r keywords}
keywords <- radtools::dicom_all_valid_header_keywords()
head(keywords, 10)
```

#### Search the DICOM standard for attribute names and keywords matching a given string:

```{r dicom_std_search}
radtools::dicom_search_header_names("manufacturer")
radtools::dicom_search_header_keywords("manufacturer")
```

## oro.dicom

Functionality not provided in oro.dicom


# Session info

```{r session_info}
sessionInfo()
```
